---
title: "Excercise 7"
output: Wage Discrimination
---


```{r}
library(haven)
library(ggplot2)
library(stargazer)
library(xtable)
library(tidyr)
library(dplyr)
library(car)
library(strucchange)
```

```{r}
data <- read_dta("Assignment3/Data/chap5-cps.dta")
data$wage <- exp(data$lnwage)
cps78 <- subset(data,year==1978)
cps85 <- subset(data,year==1985)

Male <- subset(cps78,fe==0)
Female <- subset(cps78,fe==1)
Male85 <- subset(cps85,fe==0)
Female85 <- subset(cps85,fe==1)
```

```{r}
modelI <- lm(lnwage~fe+union+hisp+nonwh+ed+ex+exsq,cps78)
linearHypothesis(modelI, "fe = 0")
stargazer(modelI,
          type = "latex",title = "Wage Regressions
          in 1978",
          dep.var.labels = "Log(Wage)",
          covariate.labels =c("Female","Union", "Hispanic", "Non-White","Education", "Expierience","Experience²"),
          digits = 3,out = "Assignment3/tables/main_models.tex")
```


```{r}
summarymale <- c(mean(Male$lnwage),mean(Male$union),mean(Male$hisp),mean(Male$nonwh),mean(Male$ed),mean(Male$ex),mean(Male$exsq))
summaryfemale <- c(mean(Female$lnwage),mean(Female$union),mean(Female$hisp),mean(Female$nonwh),mean(Female$ed),mean(Female$ex),mean(Female$exsq))

summary <- data.frame(summarymale,summaryfemale)
colnames(summary) <- c("Male","Female")
row.names(summary) <- c("LN Wage","Union","Hispanic","Non-White","ED","EX","EXSQ")
summary$MF <- summary$Male-summary$Female
```

```{r}
summary_table <- as.data.frame(summary)
summary_table <- t(summary_table)

# 1️⃣ Spaltennamen (jetzt: Male, Female, Difference)
rownames(summary_table) <- c("Male", "Female", "Difference (F - M)")

# 2️⃣ Zeilennamen setzen (die Variablen)
colnames(summary_table) <- c("LN(Wage)", "Union", "Hispanic", "Non-White", 
                             "Education", "Experience", "Experience²")

# 3️⃣ Rundung
summary_table <- round(summary_table, 3)

# 4️⃣ Stargazer-Tabelle erzeugen
stargazer(summary_table,
          summary = FALSE,
          title = "Sample Means by Gender (CPS78)",
          label = "tab:gender_means",
          digits = 3,
          out = "Assignment3/tables/gender_means2.tex",
          rownames = TRUE)
```

```{r}
modelII <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Male)
modelIII <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Female)

stargazer(modelII, modelIII,
          type = "latex",title = "Wage Regressions by Gender in 1978",
          column.labels = c("Male", "Female"),dep.var.labels = "Log(Wage)",
          covariate.labels =c("Union", "Hispanic", "Non-White","Education", "Expierience","Experience²"),
          digits = 3,out = "Assignment3/tables/gender_models.tex")

```

```{r}
modelIV <- lm(lnwage ~union+hisp+nonwh+ed+ex+exsq,data)

RSS_IV <- sum(resid(modelIV)^2)
RSS_III <- sum(resid(modelIII)^2)
RSS_II <- sum(resid(modelII)^2)

k_IV <- length(coef(modelIV))
n_III <- nobs(modelIII)
n_II <- nobs(modelII)

summary2 <- data.frame(c(RSS_IV,RSS_III,RSS_II),c(k_IV,n_III,n_II))

F_chow <- ((RSS_IV-(RSS_II+RSS_III))/k_IV)/((RSS_II+RSS_III)/(n_II+n_III-2*k_IV))
p_value <- pf(F_chow, k_IV, (n_II + n_III - 2*k_IV), lower.tail = FALSE)
```

Oxaca decomposition
Set of coefficient Beta "Rules" of the game -> this set of Charecteristic are different 

```{r}
gap <- summarymale[1] - summaryfemale[1]
bm <- coef(modelII)
bf <- coef(modelIII)

Xm <- c(1, summarymale[2:7])  
Xf <- c(1, summaryfemale[2:7])

summary3 <- data.frame(bm,bf,Xm,Xf)
explained <- sum((Xm - Xf) * bm)
unexplained <- sum(Xf * (bm - bf))

explained/gap
unexplained/gap
gap
explained + unexplained
summary3


```
Repeat for the wage discriminination between 78 85 
```{r}
modelI <- lm(lnwage~fe+union+hisp+nonwh+ed+ex+exsq,cps85)
linearHypothesis(modelI, "fe = 0")
stargazer(modelI, type = "text", title = "Wage Regression (CPS78)", digits = 3)
```
```{r}
summarymale85 <- c(mean(Male85$lnwage),mean(Male85$union),mean(Male85$hisp),mean(Male85$nonwh),mean(Male85$ed),mean(Male85$ex),mean(Male85$exsq))
summaryfemale85 <- c(mean(Female85$lnwage),mean(Female85$union),mean(Female85$hisp),mean(Female85$nonwh),mean(Female85$ed),mean(Female85$ex),mean(Female85$exsq))

summary85 <- data.frame(summarymale85,summaryfemale85)
colnames(summary85) <- c("Male","Female")
row.names(summary85) <- c("LN Wage","Union","Hispanic","Non-White","ED","EX","EXSQ")
summary85$MF <- summary85$Male-summary85$Female
```

```{r}
modelII85 <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Male85)
modelIII85 <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Female85)

stargazer(modelII85, modelIII85,
          type = "latex",title = "Wage Regressions by Gender in 1985",
          column.labels = c("Male", "Female"),dep.var.labels = "Log(Wage)",
          covariate.labels =c("Union", "Nonwhite", "Hispanic","Education", "Experience²"),
          digits = 3,out = "Assignment3/tables/gender_models85.tex")

```

```{r}
modelIV85 <- lm(lnwage ~union+hisp+nonwh+ed+ex+exsq,data)

RSS_IV85 <- sum(resid(modelIV85)^2)
RSS_III85 <- sum(resid(modelIII85)^2)
RSS_II85 <- sum(resid(modelII85)^2)

k_IV85 <- length(coef(modelIV85))
n_III85 <- nobs(modelIII85)
n_II85 <- nobs(modelII85)

summary285 <- data.frame(c(RSS_IV85,RSS_III85,RSS_II85),c(k_IV85,n_III85,n_II85))

F_chow85 <- ((RSS_IV85-(RSS_II85+RSS_III85))/k_IV85)/((RSS_II85+RSS_III85)/(n_II85+n_III85-2*k_IV85))
p_value85 <- pf(F_chow85, k_IV85, (n_II85 + n_III85 - 2*k_IV85), lower.tail = FALSE)
```

Oxaca decomposition
Set of coefficient Beta "Rules" of the game -> this set of Charecteristic are different 

```{r}
gap85 <- summarymale85[1] - summaryfemale85[1]
bm85 <- coef(modelII85)
bf85 <- coef(modelIII85)

Xm85 <- c(1, summarymale85[2:7])  
Xf85 <- c(1, summaryfemale85[2:7])

summary385 <- data.frame(bm85,bf85,Xm85,Xf85)
explained85 <- sum((Xm85 - Xf85) * bm85)
unexplained85 <- sum(Xf85 * (bm85 - bf85))

explained85/gap85
unexplained85/gap85
gap85
explained85 + unexplained85
summary385
```
Okay hier bekommen wir negative Ergebnisse das Bedeutet, das die Frauen eine bessere ausstattung (Fuer Lohn relevante merkmale) hatten und dennoch weniger verdient haben also sehen wir eine erhoete diskriminierung im vergleich zu dem was


```{r}
oaxaca_df <- data.frame(
  year = c("1978", "1985"),
  explained   = c(explained, explained85),  
  unexplained = c(unexplained, unexplained85))

oaxaca_long <- oaxaca_df %>%
  pivot_longer(
    cols = c(explained, unexplained),
    names_to = "component",
    values_to = "value"
  )

p <- ggplot(oaxaca_long, aes(x = year, y = value, fill = component)) +
  geom_bar(stat = "identity", color = "black",position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_text(aes(label = paste0(round(value, 2))),
            position = position_dodge(width = 0.8),
            vjust = -0.3,
            size = 3.5) +
  scale_fill_manual(
    values = c("explained" = "gray", "unexplained" = "red"),
    labels = c("Explained (endowments)", "Unexplained (discrimination)"),
    name = ""
  ) +
  labs(
    y = "Log points wage gap"
  ) +
  theme_minimal(base_size = 13) +
 theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.margin = margin(10, 10, 10, 10)
  )

print(p)
ggsave(filename = "Assignment3/Figures/Oaxaca-Blinder.png",plot = p,width = 7,height = 4.5,units = "in",dpi = 300)     
```
```{r}
oaxaca_share <- oaxaca_df %>%
  mutate(
    total = explained + unexplained,
    explained = explained / total,
    unexplained = unexplained / total)

oaxaca_df$type <- "Total (log points)"
oaxaca_share$type <- "Share (%)"

oaxaca_long <- bind_rows(oaxaca_df, oaxaca_share) %>%
  pivot_longer(cols = c(explained, unexplained),
               names_to = "component",
               values_to = "value")

# --- Convert shares to % and keep only the share part ---
oaxaca_long2 <- oaxaca_long %>%
  mutate(value = ifelse(type == "Share (%)", value * 100, value)) %>%
  filter(type == "Share (%)")   # 👈 only keep the share panel

```
Plot sahres of how much of the wage gap can be explained cannot be explained 
```{r}
pp <- ggplot(oaxaca_long2, aes(x = year, y = value, fill = component)) +
  geom_bar(stat = "identity", color = "black",position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_text(aes(label = paste0(round(value, 2),"%")),
            position = position_dodge(width = 0.8),
            vjust = -0.3,
            size = 3.5) +
  scale_fill_manual(
    values = c("explained" = "gray", "unexplained" = "red"),
    labels = c("Explained (endowments)", "Unexplained (discrimination)"),
    name = ""
  ) +
  labs(
    y = "Share of total wage gap", x=NULL
  ) +
  theme_minimal(base_size = 13) +
 theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.margin = margin(10, 10, 10, 10)
  )

ggsave(filename = "Assignment3/Figures/Oaxaca-Blindershare.png",plot = pp,width = 7,height = 4.5,units = "in",dpi = 300)  
print(pp)
```
```{r}
# Example for 1978
Female_married <- subset(Female, marr == 1)
Female_single  <- subset(Female, marr == 0)

summaryfemale_married <- c(mean(Female_married$lnwage),mean(Female_married$union),mean(Female_married$hisp),mean(Female_married$nonwh),mean(Female_married$ed),mean(Female_married$ex),mean(Female_married$exsq))
summaryfemale_single <- c(mean(Female_single$lnwage),mean(Female_single$union),mean(Female_single$hisp),mean(Female_single$nonwh),mean(Female_single$ed),mean(Female_single$ex),mean(Female_single$exsq))

modelIIIFmarr <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Female_married)
modelIIIFsingle <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Female_single)
```


```{r}
gapmsingle <- summarymale[1] - summaryfemale_single[1]
bf_single <- coef(modelIIIFsingle)
Xf_single <- c(1, summaryfemale_single[2:7])

summary3msingle <- data.frame(bm,bf_single,Xm,Xf_single)
explainedmsingle78 <- sum((Xm - Xf_single) * bm)
unexplainedmsingle78 <- sum(Xf_single * (bm - bf_single))

explainedmsingle78/gapmsingle
unexplainedmsingle78/gapmsingle
gapmsingle
explainedmsingle78 + unexplainedmsingle78
summary3msingle

gapmmarried <- summarymale[1] - summaryfemale_married[1]
bf_married <- coef(modelIIIFmarr)
Xf_married <- c(1, summaryfemale_married[2:7])

summary3mmarried <- data.frame(bm,bf_married,Xm,Xf_married)
explainedmmarried78 <- sum((Xm - Xf_married) * bm)
unexplainedmmarried78 <- sum(Xf_married * (bm - bf_married))

explainedmmarried78/gapmmarried
unexplainedmmarried78/gapmmarried
gapmmarried
explainedmmarried78 + unexplainedmmarried78
summary3mmarried
```
```{r}
oaxaca_dfmarr78 <- data.frame(
  year = c("Single", "Married"),
  explained78   = c(explainedmsingle78, explainedmmarried78),  
  unexplained78 = c(unexplainedmsingle78, unexplainedmmarried78))

oaxaca_longmarr <- oaxaca_dfmarr %>%
  pivot_longer(
    cols = c(explained, unexplained),
    names_to = "component",
    values_to = "value"
  )

pmarr78 <- ggplot(oaxaca_longmarr, aes(x = year, y = value, fill = component)) +
  geom_bar(stat = "identity", color = "black",position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_text(aes(label = paste0(round(value, 2))),
            position = position_dodge(width = 0.8),
            vjust = -0.3,
            size = 3.5) +
  scale_fill_manual(
    values = c("explained" = "gray", "unexplained" = "red"),
    labels = c("Explained (endowments)", "Unexplained (discrimination)"),
    name = ""
  ) +
  scale_y_continuous(limits = c(0, 0.5))+
  labs(
    y = "Log points wage gap",
    x = "1978"
  ) +
  theme_minimal(base_size = 13) +
 theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.margin = margin(10, 10, 10, 10)
  )
pmarr78
```

```{r}
Female_married85 <- subset(Female85, marr == 1)
Female_single85  <- subset(Female85, marr == 0)

summaryfemale_married85 <- c(mean(Female_married85$lnwage),mean(Female_married85$union),mean(Female_married85$hisp),mean(Female_married85$nonwh),mean(Female_married85$ed),mean(Female_married85$ex),mean(Female_married85$exsq))
summaryfemale_single85 <- c(mean(Female_single85$lnwage),mean(Female_single85$union),mean(Female_single85$hisp),mean(Female_single85$nonwh),mean(Female_single85$ed),mean(Female_single85$ex),mean(Female_single85$exsq))

modelIIIFmarr85 <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Female_married85)
modelIIIFsingle85 <- lm(lnwage~union+hisp+nonwh+ed+ex+exsq,Female_single85)
```

```{r}
gapmsingle85 <- summarymale85[1] - summaryfemale_single85[1]
bf_single85 <- coef(modelIIIFsingle85)
Xf_single85 <- c(1, summaryfemale_single85[2:7])

summary3msingle85 <- data.frame(bm85,bf_single85,Xm85,Xf_single85)
explainedmsingle85 <- sum((Xm85 - Xf_single85) * bm85)
unexplainedmsingle85 <- sum(Xf_single85 * (bm85 - bf_single85))

explainedmsingle85/gapmsingle85
unexplainedmsingle85/gapmsingle85
gapmsingle85
explainedmsingle85 + unexplainedmsingle85
summary3msingle85

gapmmarried85 <- summarymale85[1] - summaryfemale_married85[1]
bf_married85 <- coef(modelIIIFmarr85)
Xf_married85 <- c(1, summaryfemale_married85[2:7])

summary3mmarried85 <- data.frame(bm85,bf_married85,Xm85,Xf_married85)
explainedmmarried85 <- sum((Xm85 - Xf_married85) * bm85)
unexplainedmmarried85 <- sum(Xf_married85 * (bm85 - bf_married85))

explainedmmarried85/gapmmarried85
unexplainedmmarried85/gapmmarried85
gapmmarried85
explainedmmarried85 + unexplainedmmarried85
summary3mmarried85
```

```{r}
oaxaca_dfmarr85 <- data.frame(
  year = c("Single", "Married"),
  explained   = c(explainedmsingle85, explainedmmarried85),  
  unexplained = c(unexplainedmsingle85, unexplainedmmarried85))

oaxaca_longmarr85 <- oaxaca_dfmarr85 %>%
  pivot_longer(
    cols = c(explained, unexplained),
    names_to = "component",
    values_to = "value"
  )

pmarr85 <- ggplot(oaxaca_longmarr85, aes(x = year, y = value, fill = component)) +
  geom_bar(stat = "identity", color = "black",position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_text(aes(label = paste0(round(value, 2))),
            position = position_dodge(width = 0.8),
            vjust = -0.3,
            size = 3.5) +scale_y_continuous(limits = c(-0.1, 0.5w))+
  scale_fill_manual(
    values = c("explained" = "gray", "unexplained" = "red"),
    labels = c("Explained (endowments)", "Unexplained (discrimination)"),
    name = ""
  ) +
  labs(
    y = "Log points wage gap",
    x = "1985"
  ) +
  theme_minimal(base_size = 13) +
 theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.margin = margin(10, 10, 10, 10)
  )
pmarr85
```


```{r}
ggsave(filename = "Assignment3/Figures/marriedsingle78.png",plot = pmarr78 ,width = 7,height = 4.5,units = "in",dpi = 300)
ggsave(filename = "Assignment3/Figures/marriedsingle85.png",plot = pmarr85,width  = 7,height = 4.5,units = "in",dpi = 300)
```

